---
title: "Assignment No. 1"
output:
  html_document:
    df_print: paged
---
1. Explore the overall structure of the dataset using “str” and “summary” functions.
```{r}
realtor <- read.csv("C:\\Users\\Precision T3620\\Downloads\\realtor-1.csv")

str(realtor)

summary(realtor)
```

2. Specify the type of each variable
```{r}
cat("status; Catagorical, nominal\n
    price; numeric, continuous\n
    bed; numeric, discrete\n
    bath; numeric, discrete\n
    acre_lot; numeric, continuous\n
    full_address; Catagorical, nominal\n
    street; Catagorical, nominal\n
    city; Catagorical, nominal\n
    state; Catagorical, nominal\n
    zip_code; numeric, discrete\n
    house_size; numeric, continuous\n
    sold_date; Catagorical, nominal\n")
```

3. Are there any duplicate observations in the data? If so, remove them. You can use “duplicated” or “unique” functions to answer this question.
```{r}

duplicates <- duplicated(realtor)

num_duplicates <- sum(duplicates)
cat('yes there are many duplicate values, i.e\n')
print(num_duplicates)

realtor_unique <- realtor[!duplicates, ]
print(realtor_unique)

```


4. Does any of the variables have missing values? Which ones?
```{r}
missing_values <- colSums(is.na(realtor))

variables_with_missing <- names(missing_values[missing_values > 0])
print("missing values in Dataset")
print(variables_with_missing)

print("Count of missing values in price column")
sum(is.na(realtor$price))
print("Count of missing values in price bed")
sum(is.na(realtor$bed))
print("Count of missing values in price bath")
sum(is.na(realtor$bath))
print("Count of missing values in price acre_lot")
sum(is.na(realtor$acre_lot))
print("Count of missing values in price zip_code")
sum(is.na(realtor$zip_code))
print("Count of missing values in price house_size")
sum(is.na(realtor$house_size))

```
5. Remove all houses with price less than or equal to 50K
```{r}

cleaned_data <- realtor[realtor$price > 50000, ]
print(cleaned_data)

```

6. The price variable appears to have some extreme values. Remove the outliers in the “price” variable using the IQR method. IQR and quantile functions throw error if you have NAs in your variable. Use na.rm=TRUE option inside IQR and quantile methods to ignore the missing price values
```{r}
q1 <- quantile(realtor$price, na.rm = TRUE)[2]  
q3 <- quantile(realtor$price, na.rm = TRUE)[4]  
iqr <- IQR(realtor$price, na.rm = TRUE)  

lower_bound <- q1 - 1.5 * iqr
upper_bound <- q3 + 1.5 * iqr

cleaned_data <- realtor[realtor$price >= lower_bound & realtor$price <= upper_bound, ]
summary(cleaned_data)

```
7. Draw a histogram and boxplot of the price. What can you say about the shape of price
variable? Is the price variable positively skewed, symmetric, or negatively skewed?

```{r}
install.packages("gridExtra")
library(ggplot2)

histogram <- ggplot(cleaned_data, aes(x = price)) +
  geom_histogram(binwidth = 10000, fill = "skyblue", color = "black", aes(y = ..density..), alpha = 0.7) +
  geom_density(alpha = 0.2, fill = "orange") +
  labs(title = "Histogram of Price", x = "Price", y = "Density")

boxplot <- ggplot(cleaned_data, aes(y = price)) +
  geom_boxplot(fill = "lightgreen", color = "black") +
  labs(title = "Boxplot of Price", y = "Price")

library(gridExtra)
grid.arrange(histogram, boxplot, ncol = 2)

```

```{r}

cat('Variable has a summary that shows positive skewness. The mean value is higher than the median value, indicating that there are more lower values and fewer high values. The histogram of the variable will likely show a long tail on the right side, with most of the data concentrated at the lower end')
```
8. what percentage of the observations are missing for the price variable?
```{r}

total_observations <- nrow(realtor)
missing_price <- sum(is.na(realtor$price))
percentage_missing <- (missing_price / total_observations) * 100
print(percentage_missing)

```
9. Use as.Date method to convert the sold_date variable from character string to date/time type.
Then from this date/time object create two new attributes (sold_year) and (sold_month) to store the year and
month that the house was sold
```{r}
realtor$sold_date <- as.Date(realtor$sold_date, format = "%Y-%m-%d")
realtor$sold_year <- format(realtor$sold_date, "%Y")
realtor$sold_month <- format(realtor$sold_date, "%m")
str(realtor)
```
10. convert the “state” attribute to factor then take a summary to see how many observations are
there for each state. Remove states with only one observation from the data
```{r}
realtor$state <- factor(realtor$state)
state_summary <- table(realtor$state)
print(state_summary)
states_to_remove <- names(state_summary[state_summary == 1])

realtor <- realtor[!(realtor$state %in% states_to_remove), ]

state_summary_after_removal <- table(realtor$state)
print(state_summary_after_removal)

```
11. Is there a statistically significant difference between the average house price for different states?
Use appropriate plot and statistic test to answer this question
```{r}
boxplot(realtor$price ~ realtor$state, data = realtor, main = "Boxplot of House Prices by State", xlab = "State", ylab = "Price")


```
```{r}
library(ggplot2)
library(stats)


anova_result <- aov(realtor$price ~ realtor$state, data = realtor)
print(summary(anova_result))

ggplot(realtor, aes(x = realtor$state, y = realtor$price)) +
  geom_boxplot() +
  labs(title = "Average House Prices by State", x = "State", y = "Price")

p.value <- summary(anova_result)$`Pr(>F)`[1]

if (isTRUE( p.value < 0.05)) {
  print("There is a statistically significant difference between the average house prices for the differentdifferent states.")
} else {
  print("There is not a statistically significant difference between the average house prices for the different states.")
}

```



```{r}
realtor$price <- as.numeric(realtor$price)
realtor$sold_year <- as.numeric(realtor$sold_year)
realtor$house_size <- as.numeric(realtor$house_size)
realtor$bed <- as.numeric(realtor$bed)
realtor$bath <- as.numeric(realtor$bath)


correlation_sold_year <- cor(realtor$price, realtor$sold_year, method = "spearman", use = "pairwise.complete.obs")


correlation_house_size <- cor(realtor$price, realtor$house_size, method = "spearman", use = "pairwise.complete.obs")


correlation_bed <- cor(realtor$price, realtor$bed, method = "spearman", use = "pairwise.complete.obs")


correlation_bath <- cor(realtor$price, realtor$bath, method = "spearman", use = "pairwise.complete.obs")


cat("The Spearman rank correlation coefficient between house_price and sold_year is:", correlation_sold_year, "\n")
cat("The Spearman rank correlation coefficient between house_price and house_size is:", correlation_house_size, "\n")
cat("The Spearman rank correlation coefficient between house_price and bed is:", correlation_bed, "\n")
cat("The Spearman rank correlation coefficient between house_price and bath is:", correlation_bath,"\n")

```

```{r}


cat("Correlation coefficient between house_price and sold_year is approximately -0.034. This indicates a very weak negative monotonic relationship between house price and sold year. In other words, there is a slight tendency for house prices to decrease slightly as the year of sale increases. \n")

cat("Correlation coefficient between house_price and house_size is approximately 0.437. This indicates a moderate positive monotonic relationship between house price and house size. In other words, as the size of the house increases, there is a tendency for the price of the house to increase. \n")

cat("Correlation coefficient between house_price and bed is approximately 0.244. This indicates a weak positive monotonic relationship between house price and the number of bedrooms. In other words, as the number of bedrooms increases, there is a tendency for the price of the house to increase \n")

cat("Correlation coefficient between house_price and bath is approximately 0.490. This indicates a moderate to strong positive monotonic relationship between house price and the number of bathrooms. In other words, as the number of bathrooms increases, there is a tendency for the price of the house to increase.  \n")



```


Problem2 — Exploring Heart Disease Dataset
1.
```{r}
file1 <- read.csv("D:\\Machine Learning\\Assignment 1\\processed.cleveland.data", header = FALSE, na.strings = "?")


file2 <- read.csv("D:\\Machine Learning\\Assignment 1\\processed.hungarian.data", header = FALSE, na.strings = "?")


file3 <- read.csv("D:\\Machine Learning\\Assignment 1\\processed.switzerland.data", header = FALSE, na.strings = "?")


file4 <- read.csv("D:\\Machine Learning\\Assignment 1\\processed.va.data", header = FALSE, na.strings = "?")


```

2.
```{r}

HeartDisease <- rbind(file1, file2, file3, file4)
colnames(HeartDisease) <- c("age", "sex", "cp", "trestbps", "chol", "fbs", "restecg", "thalach", "exang", "oldpeak", "slope", "ca", "thal", "num")

```

3.
```{r}
str(HeartDisease)

summary(HeartDisease)

missing_rows <- sum(!complete.cases(HeartDisease))
total_rows <- nrow(HeartDisease)
percentage_missing <- (missing_rows / total_rows) * 100
cat("The missing percentage is:", percentage_missing, "\n")


```

```{r}
cat("age; numeric, continuous\n
    sex; Catagorical, nominal\n
    cp; Catagorical, ordinal\n
    trestbps; numeric, continuous\n
    chol; numeric, continuous\n
    fbs; Catagorical, nominal\n
    restecg; Catagorical, nominal\n
    thalach; numeric, continuous\n
    exang; Catagorical, nominal\n
    oldpeak; numeric, continuous\n
    slope; Catagorical, ordinal\n
    ca; numeric, discrete\n
    thal; Catagorical, ordinal\n
    num; numeric, discrete\n
    *ca* has an interval scale because the difference between values is meaningful (e.g., 1 major vessel vs. 2 major vessels).*thal* is more appropriately treated as an ordinal scale, where the values represent different levels or types of thalassemia.*num* has a nominal scale, where the values represent different categories without any inherent order.")
```
4. Convert all categorical variables to “factor” using factor function ( set the “labels” option to
give meaningful names/labels to each level)
```{r}
HeartDisease$sex <- factor(HeartDisease$sex, levels = c(0, 1), labels = c("Female", "Male"))

HeartDisease$cp <- factor(HeartDisease$cp, levels = c(1, 2, 3, 4), labels = c("Typical Angina", "Atypical Angina", "Non-anginal Pain", "Asymptomatic"))

HeartDisease$fbs <- factor(HeartDisease$fbs, levels = c(0, 1), labels = c("<= 120 mg/dl", "> 120 mg/dl"))

HeartDisease$restecg <- factor(HeartDisease$restecg, levels = c(0, 1, 2), labels = c("Normal", "ST-T Wave Abnormality", "Left Ventricular Hypertrophy"))

HeartDisease$exang <- factor(HeartDisease$exang, levels = c(0, 1), labels = c("No", "Yes"))

HeartDisease$slope <- factor(HeartDisease$slope, levels = c(1, 2, 3), labels = c("Upsloping", "Flat", "Downsloping"))

HeartDisease$thal <- factor(HeartDisease$thal, levels = c(3, 6, 7), labels = c("Normal", "Fixed Defect", "Reversible Defect"))

# Took assistance with ChatGPT
str(HeartDisease)

```


```{r}
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

median_age <- median(HeartDisease$age, na.rm = TRUE)
mode_age <- Mode(HeartDisease$age)

print(paste("Median age:", median_age))
print(paste("Mode age:", mode_age))


```

```{r}

HeartDisease$diagnosis <- factor(ifelse(HeartDisease$num == 0, "No", "Yes"), levels = c("No", "Yes"))
HeartDisease <- HeartDisease[, -14]
str(HeartDisease)

```

```{r}
ggplot(HeartDisease, aes(x = HeartDisease$diagnosis, y = HeartDisease$age, fill = diagnosis)) +
  geom_boxplot() +
  labs(title = "Box Plot of Age by Diagnosis",
       x = "Diagnosis", y = "Age") +
  theme_minimal()

ggplot(HeartDisease, aes(x = HeartDisease$diagnosis, y = HeartDisease$trestbps, fill = diagnosis)) +
  geom_boxplot() +
  labs(title = "Box Plot of trestbps by Diagnosis",
       x = "Diagnosis", y = "Resting Blood Pressure") +
  theme_minimal()

ggplot(HeartDisease, aes(x = diagnosis, y = chol, fill = diagnosis)) +
  geom_boxplot() +
  labs(title = "Box Plot chol by Diagnosis",
       x = "Diagnosis", y = "Serum Cholesterol") +
  theme_minimal()

ggplot(HeartDisease, aes(x = diagnosis, y = thalach, fill = diagnosis)) +
  geom_boxplot() +
  labs(title = "Box Plot of thalach by Diagnosis",
       x = "Diagnosis", y = "Maximum Heart Rate Achieved") +
  theme_minimal()

ggplot(HeartDisease, aes(x = diagnosis, y = oldpeak, fill = diagnosis)) +
  geom_boxplot() +
  labs(title = "Box Plot of oldpeak by Diagnosis",
       x = "Diagnosis", y = "oldpeak") +
  theme_minimal()

```


```{r}

library(ggplot2)
library(dplyr)

test_and_plot_numeric_t_test <- function(data, numeric_variable, categorical_variable) {
  t_test <- t.test(data[[numeric_variable]] ~ data[[categorical_variable]])
  print(paste("T-test p-value for", numeric_variable, "and", categorical_variable, ":", format(t_test$p.value, scientific = FALSE)))
  ggplot(data, aes(x = diagnosis, y = .data[[numeric_variable]], fill = .data[[categorical_variable]])) +
    geom_boxplot(position = position_dodge()) +
    labs(title = paste("Side-by-Side Box Plot of", numeric_variable, "by", categorical_variable, "and Diagnosis"),
         x = "Diagnosis", y = numeric_variable) +
    theme_minimal()
}


test_and_plot_numeric_t_test(HeartDisease, "age", "diagnosis")
test_and_plot_numeric_t_test(HeartDisease, "trestbps", "diagnosis")
test_and_plot_numeric_t_test(HeartDisease, "chol", "diagnosis")
test_and_plot_numeric_t_test(HeartDisease, "thalach", "diagnosis")
test_and_plot_numeric_t_test(HeartDisease, "oldpeak", "diagnosis")

```

```{r}
library(ggplot2)
library(dplyr)

test_and_plot_categorical <- function(data, variable) {
  contingency_table <- table(data$diagnosis, data[[variable]])
  chi_squared_test <- chisq.test(contingency_table)
  print(paste("Chi-squared test p-value for", variable, ":", format(chi_squared_test$p.value, scientific = FALSE)))
  mosaicplot(contingency_table, main = paste("Mosaic Plot of", variable, "by Diagnosis"), color = TRUE)
}

test_and_plot_categorical(HeartDisease, "sex") 
test_and_plot_categorical(HeartDisease, "fbs")
test_and_plot_categorical(HeartDisease, "restecg")
test_and_plot_categorical(HeartDisease, "exang")

```

```{r}
library(ggplot2)
library(vcd)

chi_squared_test <- chisq.test(HeartDisease$diagnosis, HeartDisease$cp)

print(paste("Chi-squared test p-value:", format(chi_squared_test$p.value, scientific = FALSE)))

mosaicplot(table(HeartDisease$diagnosis, HeartDisease$cp),
           main = "Mosaic Plot of Chest Pain Type by Diagnosis",
           color = TRUE)


chi_squared_test_slope <- chisq.test(HeartDisease$diagnosis, HeartDisease$slope)

print(paste("Chi-squared test p-value for slope:", format(chi_squared_test_slope$p.value, scientific = FALSE)))

mosaicplot(table(HeartDisease$diagnosis, HeartDisease$slope),
           main = "Mosaic Plot of Slope by Diagnosis",
           color = TRUE)

chi_squared_test_thal <- chisq.test(HeartDisease$diagnosis, HeartDisease$thal)

print(paste("Chi-squared test p-value for thal:", format(chi_squared_test_thal$p.value, scientific = FALSE)))

mosaicplot(table(HeartDisease$diagnosis, HeartDisease$thal),
           main = "Mosaic Plot of Thal by Diagnosis",
           color = TRUE)


```

```{r}
library(ggplot2)

anova_result <- aov(ca ~ diagnosis, data = HeartDisease)

print(summary(anova_result))

ggplot(HeartDisease, aes(x = diagnosis, y = ca, fill = diagnosis)) +
  geom_boxplot() +
  labs(title = "Side-by-Side Box Plot of Number of Major Vessels Colored by Diagnosis",
       x = "Diagnosis",
       y = "Number of Major Vessels Colored") +
  theme_minimal()


```
```{r}
library(ggplot2)

t_test_result <- t.test(ca ~ diagnosis, data = HeartDisease)

print(paste("T-test p-value:", format(t_test_result$p.value, scientific = FALSE)))

ggplot(HeartDisease, aes(x = diagnosis, y = ca, fill = diagnosis)) +
  geom_boxplot() +
  labs(title = "Side-by-Side Box Plot of Number of Major Vessels Colored by Diagnosis",
       x = "Diagnosis",
       y = "Number of Major Vessels Colored") +
  theme_minimal()


```
